---
title: "Homework 3"
author: "Owen Wright"
date: "2026-05-28"
format:
  pdf:
    toc: true
execute:
  warning: false
  message: false
editor: visual
---

```{r}
#Load the tidyverse package
library(tidyverse)

# Load the here package so file paths follow correctly
library(here)

#Read in the kelp data and store as an object called kelp
kelp <- read_csv(here("data", "temp-kelp.csv"))

#Read in personal data and store as an object called personal_data
personal_data <- read_csv(here("data", "personaldata2.csv"))

```

# Problem 1. **Giant kelp fronds**

## a. An appropriate test

In order to determine the strength of the relationship between temperature and giant kelp frond elongation rate, one may employ a Pearson correlation test, which measures how strong the relationship is and whether it is positive or negative. Additionally, one may utilize a simple linear regression test to determine whether ocean temperature predicts kelp elongation rate and what the growth rate is for each 1°C increase in temperature. Both tests are applicable because both variables measured are continuous.

## b. **Create a visualization**

```{r}
#Create a scatterplot to display the relationship between temperature and
#giant kelp frond elongation rate.
ggplot(
  data = kelp,
  mapping = aes(x = temp_c,
                y = kelp_elong)) +

#Add data for each kelp observation in the data set and change color
geom_point(color = "darkgreen",
          size = 2,
          alpha = 0.75) +
  
#Relabel the both axis
labs(x = "Ocean temperature (°C)",
     y = "Giant kelp frond elongation rate (cm/day)") +
  
#Change theme
theme_light()

```

# 
